Generating exact lattices in a WFST framework

### 2011-09-26 misc/papers/icassp12_lat/paper.tex
More edits to paper.. saved some space.
git-svn-id: https://svn.code.sf.net/p/kaldi/code/trunk@519 5e6a8d80-dfce-4ca6-a32a-6e07a63d50c8

\makeatletter
\def\name#1{\gdef\@name{#1\\}}
\makeatother
\name{ Daniel Povey$^1$, Mirko Hannemann$^{1,2}$,  \\
  {Gilles Boulianne}$^3$, {Luk\'{a}\v{s} Burget}$^4$, {Arnab Ghoshal}$^5$, {Milos Janda}$^2$, {Stefan Kombrink}$^2$, \\
  {Petr Motl\'{i}\v{c}ek}$^6$, {Yanmin Qian}$^7$, {Ngoc Thang Vu}$^8$, {Korbinian Reidhammer}$^9$, {Karel Vesel\'{y}}$^2$
    \thanks{Thanks here.. remember Sanjeev.}}
%\makeatletter
%\def\name#1{\gdef\@name{#1\\}}
%\makeatother

@@

\begin{abstract}
We describe a lattice generation method that is exact, i.e. it satisfies all the 
natural properties we would want from a lattice of alternative transcriptions of
an utterance.  This method is also quite efficient; it does not 
introduce substantial overhead above one-best decoding.  Our method is
most directly applicable when using WFST decoders where the WFST is
expanded down to the HMM-state level.  It outputs
lattices that include state-level alignments as well as word labels.
The general idea is to create a state-level lattice during decoding, and to do
a special form of determinization that retains only the best-scoring path
for each word sequence.
\end{abstract}

@@

In Section~\ref{sec:wfst} we give a Weighted Finite State Transducer
(WFST) interpretation of the speech-recognition decoding problem, in order
to introduce notation for the rest of the paper.  In Section~\ref{sec:lattices}
we define the lattice generation problem, and review previous work.
In Section~\ref{sec:overview} we give an overview of our method,
and in Section~\ref{sec:details} we summarize some aspects of a determinization
algorithm that we use in our method.  In Section~\ref{sec:exp} we give
experimental results, and in Section~\ref{sec:conc} we conclude.

@@

\section{The lattice generation problem, and previous work}
\label{sec:lattices}

\subsection{Lattices, and the lattice generation problem}

There is no generally accepted single definition of a lattice.  In~\cite{efficient_general}
and~\cite{sak2010fly}, it is defined as a labeled, weighted, directed acyclic graph
(i.e. a WFSA, with word labels).  In~\cite{ney_word_graph}, time information
is also included.  In the HTK lattice format~\cite{htkbook}, phone-level time alignments 
are also supported (along with separate language model, acoustic and pronunciation-probability 
scores), and in~\cite{saon2005anatomy}, state-level alignments are also produced.
In our work here we will be producing state-level alignments; in fact, the input-symbols
on our graph, which we call {\em transition-ids}, are slightly more fine-grained
than acoustic states and contain sufficient information to reconstruct the phone
sequence.  Hence, our lattices contain at least as much information as any prior work
we cite.

There is, as far as we know, no generally accepted problem statement
for lattice generation, but all the the authors we cited seem to
be concerned with the accuracy of the information in the lattice (e.g. that the
scores and alignments are correct) and the completeness of such information (e.g.
that no high-scoring word-sequences are missing).  The simplest
way to formalize these concerns is to express them in terms of a lattice 
pruning beam $\alpha > 0$ (interpret this as a log likelihood difference).
\begin{itemize}
  \item The lattice should have a path for every word sequence within $\alpha$ of the best-scoring one.
  \item The scores and alignments in the lattice should be accurate.
  \item The lattice should not contain duplicate paths with the same word sequence.
\end{itemize}
Actually, this definition is not completely formalized-- we have left
the second condition (accuracy of scores and alignments) a little vague.
Let the lattice be $L$.  The way we would like to state this requirement is:
\begin{itemize}
  \item For every path in $L$, the score and alignment corresponds to 
   the best-scoring path in in $B$ for the corresponding word
   sequence\footnote{Or one of the best-scoring paths, in case of a tie}.
\end{itemize}
The way we actually have to state the requirement in order to get an efficient procedure is:
\begin{itemize}
  \item For every word-sequence in $B$ within $\alpha$ of the best one, the score and alignment
    for the corresponding path in $L$ is accurate.
  \item All scores and alignments in $L$ correspond to actual paths through $B$ (but not always
   necessarily the best ones).
\end{itemize}
The issue is that we want to be able to prune $B$ before generating a lattice from it,
but doing so could cause paths not within $\alpha$ of the best one to be lost, so we have
to weaken the condition.  In fact this weakening is of very little consequence, since regardless
of pruning, any word-sequence not within $\alpha$ of the best one could be omitted altogether, 
which is the same as being assigned a cost of $\infty$).

We note that by ``word-sequence'' we mean a sequence of whatever symbols are on the
output of $\HCLG$.  In our experiments these symbols represent words, but not including
silence, which we represent via alternative paths in $L$.

\subsection{Previous lattice generation methods}

Lattice generation algorithms tend to be closely linked to a particular type of decoder,
but are often justified by the same kinds of ideas.
A common assumption underlying lattice generation methods is the {\em word-pair assumption}
of~\cite{ney_word_graph}.  This is the notion that the time boundary between a pair of words
is not affected by the identity of any earlier words.  In a decoder in which there is
a different copy of the lexical tree for each preceding word, this assumption is automatically
satisfied, and if the word-pair assumption holds, in order to generate an accurate lattice it is
sufficient to store a single Viterbi back-pointer at the word level; the entire set of
such back-pointers contains enough information to generate the lattice.  Authors who have used
this type of lattice generation method~\cite{ney_word_graph,odell_thesis} have generally
not been able to evaluate how correct the word-pair assumption is in practice, but it seems
unlikely to cause problems.  Such methods are not applicable for us anyway, as we
use a WFST based decoder in which each copy of the lexical tree does not have a unique
one-word history.

The lattice generation method described in~\cite{efficient_general} 
is applicable to decoders that use WFSTs~\cite{wfst}
expanded down to the $C$ level (i.e. $CLG$), so the input symbols represent
context-dependent phones.  In WFST based decoding networks, states normally do
not have a unique one-word history, but the authors of~\cite{efficient_general}
were able to satisfy a similar condition at the phone level.  Their method was
to store a single Viterbi back-pointer at the phone level; use this to 
create a phone-level latice; prune the resulting
lattice; project it to leave only word labels; and then convert this into a word-level
lattice via epsilon removal and determinization.
Note that the form of pruning referred to here is not the same as beam pruning as
it takes account of both the forward and backward parts of the cost.
The paper also reported experiments with a different method that did not require any
phone-pair assumption; these experiments showed that the more
efficient method that relied on the phone-pair assumption
had almost the same the lattice oracle error rate as the other method.  However, the 
experiments did not evaluate how much impact the assumption had on the accuracy of the scores, 
and this information could be important in some applications.

The lattice generation algorithm that was described in~\cite{saon2005anatomy}
is applicable to WFSTs expanded down to the $H$ level (i.e. $\HCLG$),
so the input symbols represent context-dependent states.  It keeps both score and 
state-level alignment information.  In some sense this algorithm also relies
on the word-pair assumption, but since the copies of the lexical tree in the decoding
graph do not have unique word histories, the resulting algorithm has to be quite
different.  Viterbi back-pointers at the word level are used, but the algorithm keeps track
of not just a single back-pointer in each state, but the $N$ best back-pointers for the $N$ top-scoring distinct
preceding.  Therefore, this algorithm has more in common with the sentence N-best
algorithm than with the Viterbi algorithm.  By limiting $N$ to be quite small (e.g. $N{=}5$)
the algorithm was made efficient, but this is at the cost of possibly losing word sequences
that would be within the lattice-generation beam.  In addition, this algorithm was quite
complex to implement efficiently.

@@

A fairly trivial extension of the algorithm described above is to store separately
the acoustic costs and the costs arising from $\HCLG$.  This enables us to do things
like generating output from the lattice with different acoustic scaling factors.
We refer to these two costs as the graph cost and the acoustic cost, since the cost
in $\HCLG$ is not just the language model cost but also contains components
arising from transition probabilities and pronunciation probabilities.  We implement
this by using a semiring that contains two real numbers (one for the graph and one
for the acoustic costs); it keeps track of the two costs separately, but its
$\oplus$ operation returns whichever pair has the lowest sum of costs (graph plus acoustic).

Formally, if each weight is a pair $(a,b)$, then $(a,b) \otimes (c,d) = (a{+}c, b{+}d)$,
and $(a,b) \oplus (c,d)$ is equal to $(a,b)$ if $a{+}b < c{+}d$ or if $a{+}b = c{f}+d$ and $a{-}b < c{-}d$,
and otherwise is equal to $(c,d)$.  This is equivalent to the lexicographic semiring 
of~\cite{roark2011lexicographic}, on the pair $((a{+}b),(a{-}b))$.

@@

First, let us define $Q = \inv(P)$, i.e. $Q$ is the inverted, pruned state-level lattice,
where the input symbols are the words and the output symbols are the p.d.f. labels.
We want to process $Q$ in such a way that we keep only the best path through it
for each word sequence, and get the corresponding alignment.  This is possible
by defining an appropriate semiring and then doing normal determinization.  Firstly, 
let us ignore the fact that we are keeping track of separate graph and acoustic costs, 
to avoid complicating the present discussion.

We will define a semiring in which symbol sequences are encoded into the weights.
Let a weight be a pair $(c, s)$, where $c$ is a cost and $s$ is a sequence of symbols.
We define the $\otimes$ operation as $(c, s) \otimes (c', s') = (c+c', (s,s'))$, where
$(s,s')$ is $s$ and $s'$ appended together.  We define the $\oplus$ operation so that
it returns whichever pair has the smallest weight: that is, $(c,s) \oplus (c',s')$ 
equals $(c,s)$ if $c < c'$, and $(c',s')$ if $c > c'$.  If the weights are identical,
we cannot arbitrarily return the first pair because this would not satisfy the semiring
axioms.  In this case, we return the pair with the shorter string part, and if
the lengths are the same, whichever string appears first in dictionary order.

Let $E$ be an encoding of the inverted state-level lattice $Q$ 
as described above, with the same number of states and arcs; $E$ is an acceptor, with its symbols
equal to the input symbol (word) on the corresponding arc of $Q$, and
the weights on the arcs of $E$ containing both the the weight and the output symbol (p.d.f.)
on the corresponding arcs of $Q$.  Let $D = \mathrm{det}(\mathrm{rmeps}(E))$.

In the case we are interested in, determinization
will always succeed because $E$ is acyclic (as long as the original decoding
graph $\HCLG$ has no $\epsilon$-input cycles).  Because $D$ is deterministic
and $\epsilon$-free, it has only one path for each word sequence. 
Determinization preserves equivalence,
and equivalence is defined in such a way that the $\oplus$-sum of the
weights of all the paths through $E$ with a particular word-sequence, must be the same
as the weight of the corresponding path through $D$ with that word-sequence.
It is clear from the definition of $\oplus$ that this path through
$D$ has the cost and alignment of the lowest-cost path through $E$ that has the
same word-sequence on it.

@@

We implemented $\epsilon$ removal and determinization as a single algorithm 
because $\epsilon$-removal using the traditional approach would greatly
increase the size of the state-level lattice (this is mentioned 
in~\cite{efficient_general}).  Our algorithm uses data-structures
specialized for the particular type of weight we are using.  The issue
is that the determinization process often has to append a single symbol to
a string of symbols (for the state alignments), and the easiest way to
do this in ``generic'' code would involve copying the whole sequence each time.
Instead we use a data structure that enables this to be done in linear
time (it involves a hash table).

@@

Conclusions here

>>>>

\makeatletter
\def\name#1{\gdef\@name{#1\\}}
\makeatother

\name{ \em Daniel Povey$^1$, Mirko Hannemann$^{1,2}$,  \\
  \em {Gilles Boulianne}$^3$, {Luk\'{a}\v{s} Burget}$^4$, {Arnab Ghoshal}$^5$, {Milos Janda}$^2$, {Stefan Kombrink}$^2$, \\
  \em {Petr Motl\'{i}\v{c}ek}$^6$, {Yanmin Qian}$^7$, {Ngoc Thang Vu}$^8$, {Korbinian Reidhammer}$^9$, {Karel Vesel\'{y}}$^2$
    \thanks{Thanks here.. remember Sanjeev.} }

@@

\begin{abstract}
We describe a lattice generation method that is exact, i.e. it satisfies all the 
natural properties we would want from a lattice of alternative transcriptions of
an utterance.  This method does not introduce substantial overhead above one-best decoding.  Our method is
most directly applicable when using WFST decoders where the WFST is
expanded down to the HMM-state level.  It outputs
lattices that include state-level alignments as well as word labels.
The general idea is to create a state-level lattice during decoding, and to do
a special form of determinization that retains only the best-scoring path
for each word sequence.
\end{abstract}

@@

In Section~\ref{sec:wfst} we give a Weighted Finite State Transducer
(WFST) interpretation of the speech-recognition decoding problem, in order
to introduce notation for the rest of the paper.  In Section~\ref{sec:lattices}
we define the lattice generation problem, and in Section~\ref{sec:previous}
we review previous work.
In Section~\ref{sec:overview} we give an overview of our method,
and in Section~\ref{sec:details} we summarize some aspects of a determinization
algorithm that we use in our method.  In Section~\ref{sec:exp} we give
experimental results, and in Section~\ref{sec:conc} we conclude.

@@

\section{The lattice generation problem}
\label{sec:lattices}

There is no generally accepted single definition of a lattice.  In~\cite{efficient_general}
and~\cite{sak2010fly}, it is defined as a labeled, weighted, directed acyclic graph
(i.e. a WFSA, with word labels).  In~\cite{ney_word_graph}, time information
is also included.  In the HTK lattice format~\cite{htkbook}, phone-level time alignments 
are also supported (along with separate language model, acoustic and pronunciation-probability 
scores), and in~\cite{saon2005anatomy}, state-level alignments are also produced.
In our work here we will be producing state-level alignments; in fact, the input-symbols
on our graph, which we call {\em transition-ids}, are slightly more fine-grained
than acoustic states and contain sufficient information to reconstruct the phone
sequence.

There is, as far as we know, no generally accepted problem statement
for lattice generation, but all the the authors we cited seem to
be concerned with the accuracy of the information in the lattice (e.g. that the
scores and alignments are correct) and the completeness of such information (e.g.
that no high-scoring word-sequences are missing).  The simplest
way to formalize these concerns is to express them in terms of a lattice 
pruning beam $\alpha > 0$ (interpret this as a log likelihood difference).
\begin{itemize}
  \item The lattice should have a path for every word sequence within $\alpha$ of the best-scoring one.
  \item The scores and alignments in the lattice should be accurate.
  \item The lattice should not contain duplicate paths with the same word sequence.
\end{itemize}
Actually, this definition is not completely formalized-- we have left
the second condition (accuracy of scores and alignments) a little vague.
Let the lattice be $L$.  The way we would like to state this requirement is:
\begin{itemize}
  \item For every path in $L$, the score and alignment corresponds to 
   the best-scoring path in in $B$ for the corresponding word
   sequence\footnote{Or one of the best-scoring paths, in case of a tie.}.
\end{itemize}
The way we actually have to state the requirement in order to get an efficient procedure is:
\begin{itemize}
  \item For every word-sequence in $B$ within $\alpha$ of the best one, the score and alignment
    for the corresponding path in $L$ is accurate.
  \item All scores and alignments in $L$ correspond to actual paths through $B$ (but not always
   necessarily the best ones).
\end{itemize}
The issue is that we want to be able to prune $B$ before generating a lattice from it,
but doing so could cause paths not within $\alpha$ of the best one to be lost, so we have
to weaken the condition.  This is no great loss, since regardless
of pruning, any word-sequence not within $\alpha$ of the best one could be omitted altogether, 
which is the same as being assigned a cost of $\infty$).
By ``word-sequence'' we mean a sequence of whatever symbols are on the
output of $\HCLG$.  In our experiments these symbols represent words, but not including
silence, which we represent via alternative paths in $L$.

\section{Previous lattice generation methods}
\label{sec:previous}

Lattice generation algorithms tend to be closely linked to particular types of decoder,
but are often justified by the same kinds of ideas.
A common assumption underlying lattice generation methods is the {\em word-pair assumption}
of~\cite{ney_word_graph}.  This is the notion that the time boundary between a pair of words
is not affected by the identity of any earlier words.  In a decoder in which there is
a different copy of the lexical tree for each preceding word, 
assuming the word-pair assumption holds, in order to generate an accurate lattice it is
sufficient to store a single Viterbi back-pointer at the word level; the entire set of
such back-pointers contains enough information to generate the lattice.  Authors who have used
this type of lattice generation method~\cite{ney_word_graph,odell_thesis} have generally
not been able to evaluate how correct the word-pair assumption is in practice, but it seems
unlikely to cause problems.  Such methods are not applicable for us anyway, as we
use a WFST based decoder in which each copy of the lexical tree does not have a unique
one-word history.

The lattice generation method described in~\cite{efficient_general} 
is applicable to decoders that use WFSTs~\cite{wfst}
expanded down to the $C$ level (i.e. $CLG$), so the input symbols represent
context-dependent phones.  In WFST based decoding networks, states normally do
not have a unique one-word history, but the authors of~\cite{efficient_general}
were able to satisfy a similar condition at the phone level.  Their method was
to store a single Viterbi back-pointer at the phone level; use this to 
create a phone-level latice; prune the resulting
lattice; project it to leave only word labels; and then remove $\epsilon$
symbols and determinize.
Note that the form of pruning referred to here is not the same as beam pruning as
it takes account of both the forward and backward parts of the cost.
The paper also reported experiments with a different method that did not require any
phone-pair assumption; these experiments showed that the more
efficient method that relied on the phone-pair assumption
had almost the same lattice oracle error rate as their more efficient method.  However, the 
experiments did not evaluate how much impact the assumption had on the accuracy of the scores, 
and this information could be important in some applications.

The lattice generation algorithm that was described in~\cite{saon2005anatomy}
is applicable to WFSTs expanded down to the $H$ level (i.e. $\HCLG$),
so the input symbols represent context-dependent states.  It keeps both scores and 
state-level alignment information.  In some sense this algorithm also relies
on the word-pair assumption, but since the copies of the lexical tree in the decoding
graph do not have unique word histories, the resulting algorithm has to be quite
different.  Viterbi back-pointers at the word level are used, but the algorithm keeps track
of not just a single back-pointer in each state, but the $N$ best back-pointers for the $N$ top-scoring distinct
preceding.  Therefore, this algorithm has more in common with the sentence N-best
algorithm than with the Viterbi algorithm.  By limiting $N$ to be quite small (e.g. $N{=}5$)
the algorithm was made efficient, but at the cost of losing word sequences
that would be within the lattice-generation beam.

@@

A fairly trivial extension of the algorithm described above is to store separately
the acoustic costs and the costs arising from $\HCLG$.  This enables us to do things
like generating output from the lattice with different acoustic scaling factors.
We refer to these two costs as the graph cost and the acoustic cost, since the cost
in $\HCLG$ is not just the language model cost but also contains components
arising from transition probabilities and pronunciation probabilities.  We implement
this by using a semiring that contains two real numbers, one for the graph and one
for the acoustic costs; it keeps track of the two costs separately, but its
$\oplus$ operation returns whichever pair has the lowest sum of costs (graph plus acoustic).

Formally, if each weight is a pair $(a,b)$, then $(a,b) \otimes (c,d) = (a{+}c, b{+}d)$,
and $(a,b) \oplus (c,d)$ is equal to $(a,b)$ if $a{+}b < c{+}d$ or if $a{+}b = c{+}d$ and $a{-}b < c{-}d$,
and otherwise is equal to $(c,d)$.  This is equivalent to the lexicographic semiring 
of~\cite{roark2011lexicographic}, on the pair $((a{+}b),(a{-}b))$.

@@

First, let us define $Q = \inv(P)$, i.e. $Q$ is the inverted, pruned state-level lattice,
where the input symbols are the words and the output symbols are the p.d.f. labels.
We want to process $Q$ in such a way that we keep only the best path through it
for each word sequence, and get the corresponding alignment.  This is possible
by defining an appropriate semiring and then doing normal determinization.  We shall
ignore the fact that we are keeping track of separate graph and acoustic costs, 
to avoid complicating the present discussion.

We will define a semiring in which symbol sequences are encoded into the weights.
Let a weight be a pair $(c, s)$, where $c$ is a cost and $s$ is a sequence of symbols.
We define the $\otimes$ operation as $(c, s) \otimes (c', s') = (c+c', (s,s'))$, where
$(s,s')$ is $s$ and $s'$ appended together.  We define the $\oplus$ operation so that
it returns whichever pair has the smallest cost: that is, $(c,s) \oplus (c',s')$ 
equals $(c,s)$ if $c < c'$, and $(c',s')$ if $c > c'$.  If the costs are identical,
we cannot arbitrarily return the first pair because this would not satisfy the semiring
axioms.  In this case, we return the pair with the shorter string part, and if
the lengths are the same, whichever string appears first in dictionary order.

Let $E$ be an encoding of the inverted state-level lattice $Q$ 
as described above, with the same number of states and arcs; $E$ is an acceptor, with its symbols
equal to the input symbol (word) on the corresponding arc of $Q$, and
the weights on the arcs of $E$ containing both the the weight and the output symbol (p.d.f.),
if any, on the corresponding arcs of $Q$.  Let $D = \mathrm{det}(\mathrm{rmeps}(E))$.
Determinization
will always succeed because $E$ is acyclic (as long as the original decoding
graph $\HCLG$ has no $\epsilon$-input cycles).  Because $D$ is deterministic
and $\epsilon$-free, it has only one path for each word sequence. 
Determinization preserves equivalence,
and equivalence is defined in such a way that the $\oplus$-sum of the
weights of all the paths through $E$ with a particular word-sequence, must be the same
as the weight of the corresponding path through $D$ with that word-sequence.
It is clear from the definition of $\oplus$ that this path through
$D$ has the cost and alignment of the lowest-cost path through $E$ that has the
same word-sequence on it.

@@

We implemented $\epsilon$ removal and determinization as a single algorithm 
because $\epsilon$-removal using the traditional approach would greatly
increase the size of the state-level lattice (this is mentioned 
in~\cite{efficient_general}).  Our algorithm uses data-structures
specialized for the particular type of weight we are using.  The issue
is that the determinization process often has to append a single symbol to
a string of symbols, and the easiest way to
do this in ``generic'' code would involve copying the whole sequence each time.
Instead we use a data structure that enables this to be done in linear
time (it involves a hash table).

@@

Conclusions here [mention does not rely on word-pair assumption]
